Reads, QC and trimming

### 2025-04-07 episodes/phylogenetics.Rmd
Fixed details on blast and phylo. Edited half of QC

```bash
iqtree2 -s rpoB.einsi.clipkit.aln -m TEST -B 1000
```

Here, we tell IQ-TREE to use the alignment `-s rpoB.einsi.clipkit.aln`, and to test among the standard substitution models which one fits best (`-m TEST`). We also tell IQ-TREE to perform 1000 ultra-fast bootstraps (`-B 1000`). We'll discuss these later.

@@

```bash
iqtree2 -s rpoB.fftns.aln -m TEST -B 1000
```

>>>>

```bash
iqtree2 -s rpoB.einsi.clipkit.aln -m MFP -B 1000
```

Here, we tell IQ-TREE to use the alignment `-s rpoB.einsi.clipkit.aln`, and to test among the standard substitution models which one fits best (`-m MFP`). We also tell IQ-TREE to perform 1000 ultra-fast bootstraps (`-B 1000`). We'll discuss these later.

@@

```bash
iqtree2 -s rpoB.fftns.aln -m MFP -B 1000
```

### 2025-04-07 episodes/quality-control.Rmd
Fixed details on blast and phylo. Edited half of QC

- Trace an outbreak of Mycobacterium, from reads to phylogenetic tree.
- How can I organize my file system for a new bioinformatics project?
- How and where can data be downloaded?
- How can I describe the quality of my data?
- How can I get rid of sequence data that doesn't meet my quality standards?

@@

Tuberculosis (TB) is an infectious disease caused by *Mycobacterium tuberculosis*. It mostly affects the lungs. An infection with *M. tuberculosis* is often asymptomatic (latent infection). Only in about 10% of the cases the latent infection progresses to an active infection during a patients lifetime, which, if untreated, leads to death in about half of the cases. The symptoms of an active TB infection include cough, fever, night sweats, weight loss etc. An active TB infection can spread. Once exposed, people often have latent TB. To identify people with latent TB, a [skin test](https://www.cdc.gov/tb/publications/factsheets/testing/skintesting.htm) can be applied.

@@

```bash
interactive -A uppmax2024-2-10 -M snowy -t 04:00:00
cd /proj/g2020004/nobackup/3MK013/<username>
pwd
```

@@

::: challenge
## Exercise

@@

```bash
mkdir molepi
mkdir molepi/docs
mkdir molepi/data
mkdir molepi/results
```
:::
:::

@@

First, we switch to the data folder to store all our data

```bash
cd molepi/data 
```

@@

```bash
wget ftp://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/000/195/955/GCF_000195955.2_ASM19595v2/GCF_000195955.2_ASM19595v2_genomic.fna.gz
```

@@

::: challenge
## Challenge: What is the size of the genome?

Find out how many nucleotides the genome has. Hint:

@@

@@

The genome has 4'411'532 bp.

@@

@@

```bash
for filename in ERR029207 ERR029206 ERR026478 ERR026474 ERR026473 ERR026481 ERR026482
do
  wget --spider ftp://ftp.sra.ebi.ac.uk/vol1/fastq/"${filename:0:6}"/"${filename}"/"${filename}"_*.fastq.gz
done
```

@@

In this example, the list is seven accession numbers of reads belonging to the genomes we are interested in. Each time the loop iterates, it will assign a file name to the variable `filename` and run the `wget` command. The first time through the loop, `$filename` is `ERR029207`. The FTP site at EBI is constructed so that fastq files are stored in a `vol1/fastq` subfolder. Read files are not directly in that subfolder, but grouped in subfolders starting with the six first characters of the accession number (`ERR029` in the first case), then in the accession number subfolder. To get to the right place, the file path is constructed with the root path (`ftp://ftp.sra.ebi.ac.uk/vol1/fastq/`), to which the first 6 characters of the accession number is added through that special variable operation (`"${filename:0:6}"`) to get to the right subfolder. Finally, the correct accession subfolder (`"${filename}"`), the file name prefix  (`"${filename}"_*.fastq.gz"`) and the suffix (`_*.fastq.gz`) are added. The `*` ensures that we get reads from both ends.

Use `{}` to wrap the variable so that .fastq.gz will not be interpreted as part of the variable name. In addition, quoting the shell variables is a good practice AND necessary if your variables have spaces in them.

@@

All the read files start with ERR and have a `fastq.gz` extension. It means they are in fastq format and compressed. Softlink the files in your `molepi/data` folder, using a nested loop. The `data/fastq` folder is three levels up:

@@

@@

We can view the first complete read in one of the files our dataset by using `head` to look at the first four lines. You will use zcat to extract text from a gzipped file:

```bash
zcat ERR029206_1.fastq.gz | head -n4 
```

@@

@@

$10^{-\frac{6}{10}} \cong 0.25 $

@@

::: callout
This mainly relates to older Solexa/Illumina data, but it's essential that you know which sequencing platform was used to generate your data, so that you can tell your quality control program which encoding to use. If you choose the wrong encoding, you run the risk of throwing away good reads or (even worse) not throwing away bad reads!
:::

>>>>

- Trace an outbreak of *Mycobacterium*, from reads to phylogenetic tree.
- How can I organize my file system for a new bioinformatics project?
- How and where can data be downloaded?
- How can I describe the quality of my data?
- How can I get rid of sequence data that doesn't meet my quality standards?

@@

Tuberculosis (TB) is an infectious disease caused by *Mycobacterium tuberculosis*. It mostly affects the lungs. An infection with *M. tuberculosis* is often asymptomatic (latent infection). Only in about 10% of the cases the latent infection progresses to an active infection during a patients lifetime, which, if untreated, leads to death in about half of the cases. The symptoms of an active TB infection include cough, fever, night sweats, weight loss etc. An active TB infection can spread. Once exposed, people often have latent TB. To identify people with latent TB, a [skin test](https://www.cdc.gov/tb/hcp/testing-diagnosis/tuberculin-skin-test.html) can be applied.

@@

@@

```bash
interactive ...
cd /proj/g2020004/nobackup/3MK013/<username>
pwd
```

:::::::::::

::: instructor

```bash
interactive -A uppmax2025-3-4 -M snowy -t 04:00:00
cd /proj/g2020004/nobackup/3MK013/<username>
pwd
```
::::::::::::::

## Exercise

@@

```bash
mkdir molepi molepi/docs molepi/data molepi/results
```

::::::::::::::

@@

First, we switch to the `data` subfolder to store all our data

::: solution 
```bash
cd molepi/data 
```
::::::::::::

@@

Find the location of the gzipped `fna` file for that reference genome via NCBI's genome, by finding the FTP site and copying the link to the `*_genomic.fna.gz` file and using that as an argument to `wget`.

::: hint

```bash
wget -h
```
:::::::::

Search for *Mycobacterium tuberculosis* on the [Datasets/Genome](https://www.ncbi.nlm.nih.gov/datasets/genome/) website. Identify the "reference genome" (at the top), click on the accession number, and find the `FTP` tab. There is many files, among which `GCF_000195955.2_ASM19595v2_genomic.fna.gz`. Right-click and copy the link.

::: instructor

```bash
wget ftp://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/000/195/955/GCF_000195955.2_ASM19595v2/GCF_000195955.2_ASM19595v2_genomic.fna.gz
```
::::::::::::::

@@

## Challenge: What is the size of the genome?

Find out how many nucleotides the genome has. There is a software called SeqKit (and a corresponding module at Uppmax) that can help.

::: hint:

@@

:::::::::

@@

The genome has 4'411'532 bp. The `stat` command of `seqkit` is helpful.

@@

::: instructor

@@

```bash
for filename in ERR029207 ERR029206 ERR026478 ERR026474 ERR026473 ERR026481 ERR026482
do
  echo -e "$filename"
  wget --spider ftp://ftp.sra.ebi.ac.uk/vol1/fastq/"${filename:0:6}"/"${filename}"/"${filename}"_*.fastq.gz
done
```

@@

In this example, the list is seven accession numbers of reads belonging to the genomes we are interested in. Each time the loop iterates, it will assign a file name to the variable `filename` and `echo` (print) the filename. In the second line, it will run the `wget` command.

The first time through the loop, `$filename` is `ERR029207`. The FTP site at EBI is constructed so that fastq files are stored in a `vol1/fastq` subfolder. Read files are not directly in that subfolder, but grouped in subfolders starting with the six first characters of the accession number (`ERR029` in the first case), then in the accession number subfolder. To get to the right place, the file path is constructed with the root path (`ftp://ftp.sra.ebi.ac.uk/vol1/fastq/`), to which the first 6 characters of the accession number is added through that special variable operation (`"${filename:0:6}"`) to get to the right subfolder. Finally, the correct accession subfolder (`"${filename}"`), the file name prefix  (`"${filename}"_*.fastq.gz"`) and the suffix (`_*.fastq.gz`) are added. The `*` ensures that we get reads from both ends.

Use `{}` to wrap the variable so that `.fastq.gz` will not be interpreted as part of the variable name. In addition, quoting the shell variables is a good practice AND necessary if your variables have spaces in them.

@@

### Link the read files

Now link the files from the `data` folder, to perform quality control.

All the read files start with ERR and have a `fastq.gz` extension. It means they are in fastq format and compressed. Softlink the files in your `molepi/data` folder, using a nested loop. The `data/fastq` folder is three levels up. For each accession number there are two `fastq.gz` files, one for each end (1 or 2).

::: hint

Softlinks are done with `ln -s`. Use the same loop backbone as above, and nest a `for` loop over the two ends of the reads (1 and 2).

::::::::

```bash
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/data
for filename in <...>
do
  for <loop over 1 2>
  do
    <softlink> ../../../data/fastq/<build the filename with $ and double quotes>.fastq.gz .
  done
done
```

::: instructor

@@

:::::::::::::

::::::::::::::

@@

We can view the first complete read in one of the files our dataset by using `head` to look at the first four lines.

You will use zcat to extract text from a gzipped file.

@@

::: hint

```bash
zcat -h
head -h
```

```bash
zcat ERR029206_1.fastq.gz | head -n4 
```
::::::::::::::

::::::::::::::

@@

$10^{-\frac{6}{10}} \cong 0.25$

@@

::: callout

This mainly relates to older Solexa/Illumina data, but it's essential that you know which sequencing platform was used to generate your data, so that you can tell your quality control program which encoding to use. If you choose the wrong encoding, you run the risk of throwing away good reads or (even worse) not throwing away bad reads!
